---
title: "Group Project Check-In Deliverable February 20, 2024"
author: "Allison Stewart, Christine Kuryla, Zander De Jesus, Alana Ferris"
date: "2024-02-19"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(foreign)
library(dplyr)
library(forcats)
library(gtsummary)
library(corrr)
library(ggridges)
library(patchwork)
library(reshape2)
library(corrplot)
library(pheatmap)
```

# Introduction Section Outline 

# Research Question 
Does exposure to phthalate alternatives affect inflammation in children as measured by alpha-1-acid glycoprotein?
We will investigate this question using NHANES data from 2017-2018 of the serum inflammatory biomarker alpha-1-acid glycoprotein and urinary phthalate alternative metabolite levels, along with other relevant data (urinary creatinine levels and sociodemographic data).

# Data Setup

### Reading in Relevant NHANES 2017-2018 Data
* NOTE: phthalate and AGP data have special sample weights we need to account for in analysis
* weighting module: https://wwwn.cdc.gov/nchs/nhanes/tutorials/Weighting.aspx
```{r message=FALSE, warning=FALSE}
# phthalates and plasticizers metabolites - urine 
# codebook, LODs: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/PHTHTE_J.htm
phthte_urine =
  read.xport("data/PHTHTE_J.XPT") %>% 
  janitor::clean_names()

# urine creatinine levels 
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/ALB_CR_J.htm
creatinine_urine =
  read.xport("data/ALB_CR_J.XPT") %>% 
  janitor::clean_names()

# Alpha-1-Acid Glycoprotein - Serum
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/SSAGP_J.htm
agp_serum =
  read.xport("data/SSAGP_J.XPT") %>% 
  janitor::clean_names()

# Demographic variables and sample weights 
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DEMO_J.htm
demo =
  read.xport("data/DEMO_J.XPT") |> 
  janitor::clean_names()
```

### Cleaning Variable Names of Phthalate Urine Data 
```{r message=FALSE, warning=FALSE}
phthte_urine =
  phthte_urine %>% 
  rename(weights = wtsb2yr) %>%
  rename_with(~ gsub("^urx", "", .), starts_with("urx"))
  
# for now, leaving the <LOD comment columns (i.e, columns started wit urd) as they are 
# 2986 obs
```

### Adjusting Phthalate Urine Data for Creatinine
```{r message=FALSE, warning=FALSE}
# merging Phthalate urine data with urinary creatinine levels and then adjusting concentrations for creatinine
# dropping albumin from creatinine dataset
creatinine_urine =
  creatinine_urine %>% 
  select(seqn, urxucr, urducrlc) #dropping urxcrs bc i dont want creatinine expressed in those units
#7936 obs

#merging creatinine data with phthalate data
phthte_creatinine =
  merge(creatinine_urine, phthte_urine, by = "seqn") %>%
  select(!starts_with("urd")) %>% #dropped <LOD columns for ease of pivoting
  pivot_longer(cols = cnp:mzp,  
               names_to = "analyte_code",
               values_to = "reported_result") #pivoted longer so that i can map in the next step

#56734 obs 

#urinary [phth] in ng/mL
# urinary [creatinine] in mg/dL so need to get this to ng/mL

# Urinary creatinine concentrations, specific gravity, and osmolality are common methods for adjusting dilution. The most widely used method is creatinine adjustment that involves dividing the analyte concentration by the creatinine concentration. Analyte results are then reported as weight of analyte per gram of creatinine (micrograms analyte per gram creatinine).

# making a new column with creatinine-adjusted [phthalate]
phthte_creatinine =
  phthte_creatinine %>% 
  mutate(ng_per_ml_creatinine = urxucr*10000) %>% #here converting mg/dL to ng/mL--> 1mg/dL = 10,000ng/mL
  mutate(adj_conc = reported_result/ng_per_ml_creatinine)

#visualizing adjusted concentrations but on the log scale bc the numbers are now so small lol ... i wonder if im just being dumb or if NHANES really doesnt report specific gravity data 
boxplot(phthte_creatinine$adj_conc, ylim = c(0, 0.08))
hist(phthte_creatinine$adj_conc, ylim = c(0, 0.02))
```

### Cleaning Variables Name of Demographic Data
```{r message=FALSE, warning=FALSE}
# drop unneeded variables from dataset 
demo <- subset(demo, select = -c(sddsrvyr, ridstatr, ridexprg, dmdmartl, dmdeduc2, dmqmiliz, dmqadfc, dmdhrgnd, dmdhragz, dmdhredz, dmdhrmaz, dmdhsedz))

demo <- demo |> 
  rename(gender = riagendr, 
         age_yrs_screen = ridageyr, 
         age_mon_screen = ridagemn, 
         race_ethn1 = ridreth1,
         race_ethn3 = ridreth3, 
         time_period = ridexmon, 
         age_mon_screen = ridagemn,
         age_mon_exam = ridexagm, 
         birth_country = dmdborn4, 
         citizenship = dmdcitzn,
         years_us = dmdyrsus, 
         education = dmdeduc3, 
         lang_pers_quest = sialang, 
         proxy_pers_quest = siaproxy, 
         interpret_pers_quest = siaintrp, 
         lang_fam_interview = fialang, 
         proxy_fam_interview = fiaproxy, 
         interpret_fam_interview = fiaintrp,
         lang_capi = mialang, 
         proxy_capi = miaproxy, 
         interpret_capi = miaintrp, 
         lang_acasi = aialanga, 
         hh_size = dmdhhsiz, 
         fam_size = dmdfmsiz, 
         hh_young_children = dmdhhsza,
         hh_children = dmdhhszb, 
         hh_adults = dmdhhsze, 
         weight_2yr_intervew = wtint2yr,
         weight_2yr_mec = wtmec2yr, 
         variance_psu = sdmvpsu, 
         variance_stratum = sdmvstra, 
         income_hh_annual = indhhin2, 
         income_family_annual = indfmin2, 
         income_poverty_ratio = indfmpir)
```

### Subset Demographic Data to Ages 3-16 
```{r message=FALSE, warning=FALSE}
demo <- demo |> 
  filter(age_yrs_screen >= 3 & age_yrs_screen <= 16)
```

# Table 1 
```{r message=FALSE, warning=FALSE}

# create table 1 with age, gender, race/ethnicity, and household income

# recode age into age group 
demo <- demo |> 
  mutate(age_group = case_when(age_yrs_screen >= 3 & age_yrs_screen < 7 ~ 1,
                                          age_yrs_screen >= 7 & age_yrs_screen < 11 ~ 2, 
                                          age_yrs_screen >= 11 & age_yrs_screen < 14 ~ 3, 
                                          age_yrs_screen >= 14 ~ 4,
                                          TRUE ~ NA_real_))

# recode household income into fewer categories 
demo <- demo |> 
  mutate(income_hh_combined = case_when(
                                      income_hh_annual == 1 ~ "1",
                                      income_hh_annual == 2 ~ "1",
                                      income_hh_annual == 3 ~ "1",
                                      income_hh_annual == 4 ~ "1",
                                      income_hh_annual == 5 ~ "2",
                                      income_hh_annual == 6 ~ "2",
                                      income_hh_annual == 7 ~ "2",
                                      income_hh_annual == 8 ~ "3",
                                      income_hh_annual == 9 ~ "3",
                                      income_hh_annual == 10 ~ "3",
                                      income_hh_annual == 14 ~ "4",
                                      income_hh_annual == 15 ~ "4",
                                      income_hh_annual %in% c(77, 99) ~ "Missing",
                                      TRUE ~ NA_character_ ))

demo$income_hh_combined <- ifelse(is.na(demo$income_hh_combined), "Missing", demo$income_hh_combined)
         
demo <- demo |> 
  mutate(income_hh_binary = case_when(income_hh_annual == 13 ~ "1",
                                      income_hh_annual == 12 ~ "2",
                                      income_hh_annual %in% c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 14,
                                                              15, 77, 99) ~ "Missing", TRUE ~ NA))

demo$income_hh_binary <- ifelse(is.na(demo$income_hh_binary), "Missing", demo$income_hh_binary)


#Recode variables of interest as factors with labels 

demo <- demo %>%
  mutate(
    age_group = factor(age_group),
    gender = factor(gender),
    race_ethn3 = factor(race_ethn3),
    income_hh_combined = factor(income_hh_combined),
    income_hh_binary = factor(income_hh_binary)
  )

demo <- demo |> 
  mutate(
    age_group = fct_recode(age_group,
                            "3-6" = "1",
                            "7-10" = "2",
                            "11-13" = "3",
                            "14-16" = "4"), 
    gender = fct_recode(gender,
                         "Female" = "1", 
                         "Male" = "2"), 
    race_ethn3 = fct_recode(race_ethn3,
                         "Mexican American" = "1", 
                         "Other Hispanic" = "2",
                         "Non-Hispanic White" = "3", 
                         "Non-Hispanic Black" = "4", 
                         "Non-Hispanic Asian" = "5", 
                         "Other" = "6", 
                         "Missing" = "7"), 
    income_hh_combined = fct_recode(income_hh_combined,
                                  "$0 to $19,999" = "1", 
                                  "$20,000 to $44,999" = "2", 
                                  "$45,000 to $74,999" = "3", 
                                  "$75,000 and Over" = "4", 
                                  "Missing" = "77", 
                                  "Missing" = "99"),
    income_hh_binary = fct_recode(income_hh_binary,
                                  "Under $20,000" = "1", 
                                  "$20,000 and Over" = "2", 
                                  "Missing" = "77", 
                                  "Missing" = "99"))

# Create table 1

# Select variables of interest
variables_of_interest <- c("age_yrs_screen", "age_group", "gender", "race_ethn3", "income_hh_combined", "income_hh_binary")


# Generate summary statistics for all variables
table_1 <- demo |> 
  select(all_of(variables_of_interest)) |> 
  tbl_summary(
    by = NULL, 
    statistic = list(all_categorical() ~ "{n} ({p}%)",
                     age_yrs_screen ~ "{median} ({sd})"),
    digits = list(all_continuous() ~ c(2, 2),
                  all_categorical() ~ c(0, 1)),
    type = list(age_yrs_screen ~ "continuous",
                age_group ~ "categorical",
                gender ~ "categorical",
                income_hh_combined ~ "categorical",
                race_ethn3 ~ "categorical",
                income_hh_binary ~ "categorical"),
    label = list(age_yrs_screen ~ "Age",
                 age_group ~ "Age Group",
                 gender ~ "Gender",
                 income_hh_combined ~ "Household Income",
                 race_ethn3 ~ "Race and Ethnicity", 
                 income_hh_binary ~ "Household Income (binary)")
    ) |> 
  modify_header(label = "**Variable**") |> 
  modify_caption("Table 1: Demographic Characteristics of Children Ages 3-16 in NHANES 2017-2018") |>
  bold_labels() |> 
  modify_table_body(
    ~.x %>%
      mutate(
        across(where(is.character), ~ifelse(.x == "9.00 (3.94)", "9 (3.94)", .x))
      )
  )

# Print the summary table
table_1

```

# Reviewing Adjusted Urinary Metabolite Data

### Summary Statistics of of Creatine-Adjusted Urinary Metabolites
```{r message=FALSE, warning=FALSE}
# metabolites matrix 
metabolite_matrix = phthte_urine |> 
  select(cnp, cop, ecp, ecpt, hibp, mbp, mc1, mcoh, mep, mhbp, mhh, mhht, mhnc, mhp, mib, mnp, moh, monp, mzp) %>% 
  na.omit()

# Calculate the summary statistics
metabolite_summary_table <- metabolite_matrix %>% summarise_all(list(
  min = ~min(., na.rm = TRUE),
  max = ~max(., na.rm = TRUE), 
  median = ~median(., na.rm = TRUE),
  sd = ~sd(., na.rm = TRUE),
  #p5 = ~quantile(., probs = 0.05, na.rm = TRUE),
  p95 = ~quantile(., probs = 0.95, na.rm = TRUE),
  percent_detect = ~sum(. > min(., na.rm = TRUE), na.rm = TRUE) / length(.) * 100
))

metabolite_summary_long <- metabolite_summary_table %>%
  pivot_longer(cols = everything(),
               names_to = "metabolite_statistic",
               values_to = "value") %>%
  separate(metabolite_statistic, into = c("metabolite", "statistic"), sep = "_", extra = "merge") %>%
  pivot_wider(names_from = "statistic", values_from = "value")

knitr::kable(metabolite_summary_long)

```

### Visualizations of Full Phthalate Metabolite Distributions, Adjusted for Creatinine
```{r message=FALSE, warning=FALSE}

#First looking at the logged10 values to see the small concentrations that Alana was mentioning
#Ordered from lowest to highest concentration distributions, can be toggled by turning off line 256
#ggplot removes 4256 rows containing non-finite values in these visualizations

metabolites_distrib_boxplot = phthte_creatinine |> 
  #mutate(log10_adj_conc = log10(adj_conc)) |> 
  mutate(analyte_code = fct_reorder(analyte_code, adj_conc)) |> 
  group_by(analyte_code) |> 
  ggplot(aes(y = adj_conc, x = analyte_code, fill = analyte_code)) + geom_boxplot() + 
  scale_y_continuous(breaks = scales::pretty_breaks(10)) + theme(legend.position = 'none') +
   labs(
    title = "Boxplot Distribution of Phthalate Metabolites found in Urine, Adjusted by Creatinine",
    x = "Metabolite Coded Name",
    y = "Adjusted Concentration (ng/mL)"
  )
  
#Completed Visualization Phthalate Concentrations (ng/mL)
metabolites_distrib_boxplot

#The untransformed concentrations are quite small, but difficult to visualize because of an extreme high outliers on the metabolite "MEP", and to a lesser extent ecpt, wanted to provide a second untransformed visualization when removing that outlier

metabolites_distrib_boxplot = phthte_creatinine |> 
  filter(adj_conc <= 0.05) |> 
  mutate(analyte_code = fct_reorder(analyte_code, adj_conc)) |> 
  group_by(analyte_code) |> 
  ggplot(aes(y = adj_conc, x = analyte_code, fill = analyte_code)) + geom_boxplot() + theme(legend.position = 'none') +
   labs(
    title = "Boxplot Distribution of Phthalate Metabolites found in Urine, Adjusted by Creatinine",
    x = "Metabolite Coded Name",
    y = "Adjusted Concentration (ng/mL)"
  )

#This visualization is not very usable, but shows the broader distribution of relatively higher concentrations of MEP and ECPT present in urine
metabolites_distrib_boxplot 


metabolites_hisgrid = phthte_creatinine |> 
  #mutate(log10_adj_conc = log10(adj_conc)) |> 
  mutate(analyte_code = fct_reorder(analyte_code, adj_conc)) |> 
  group_by(analyte_code) |> 
  ggplot(aes(x = adj_conc, fill = analyte_code)) + geom_histogram() + theme(
    legend.position = 'right') + 
  facet_wrap(. ~ analyte_code) +
  labs(title = "Histogram Distributions of Phthalate Metabolites found in Urine, Adjusted by Creatinine",
    x = "Adjusted Concentration (ng/mL)",
    y =  "Count Observations")

metabolites_hisgrid 
```

# Reviewing AGP Data 

### Summary Statistics of AGP
```{r message=FALSE, warning=FALSE}
# AGP
agp_summary_table <- agp_serum %>%
  select(ssagp) %>% 
  summarise(
    min = min(ssagp, na.rm = TRUE),
    max = max(ssagp, na.rm = TRUE),
    median = median(ssagp, na.rm = TRUE),
    mean = mean(ssagp, na.rm = TRUE),
    sd = sd(ssagp, na.rm = TRUE),
    p95 = quantile(., probs = 0.95, na.rm = TRUE),
    count = n(),
    percent_na = sum(is.na(ssagp)) / n() * 100
  )
knitr::kable(agp_summary_table)
```

### Histogram of AGP distribution 
```{r agp_hist, message=FALSE, warning=FALSE}

# histogram of AGP distribution

hist(agp_serum$ssagp,
     main = "Histogram of AGP Distribution",
     xlab = "SS AGP",
     col = "skyblue")

```


## Phthalate Metabolite Correlation Matrix
```{r message=FALSE, warning=FALSE}

# metabolites matrix 
metabolite_matrix = phthte_urine |> 
  select(cnp, cop, ecp, ecpt, hibp, mbp, mc1, mcoh, mep, mhbp, mhh, mhht, mhnc, mhp, mib, mnp, moh, monp, mzp) %>% 
  na.omit()

metabolite_corr_matrix <- cor(metabolite_matrix)

# visualize correlation matrix

library(corrplot)

corrplot(metabolite_corr_matrix, 
         method = "number",
         number.cex = 0.8,
         number.font = 2,
         number.digits = 1,
         tl.col = "black",
         addCoef.col = 'grey',
         mar = c(0, 0, 1, 0),
         main = "Metabolite Correlations")
corrplot(metabolite_corr_matrix, 
         method = "shade",
         mar = c(0, 0, 1, 0),
         tl.col = "black",
         main = "Metabolite Correlation Matrix")

```

```{r message=FALSE, warning=FALSE}

library(ggplot2)
library(reshape2)
cor_matrix_melted <- melt(as.matrix(metabolite_corr_matrix))
ggplot(cor_matrix_melted, aes(Var1, Var2, fill=value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "skyblue", mid = "lavender", high = "hotpink", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank()) +
  ggtitle("Metabolite Heatmap") +
  coord_fixed()
library(pheatmap)
pheatmap(metabolite_corr_matrix, 
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         color = colorRampPalette(c("palegreen", "turquoise", "cornflowerblue"))(100),
         main = "Metabolite Correlation Heirarchical Clustering")
```
