---
title: "data_cleaning"
author: "Alana Ferris"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(foreign)
```

# Research Question 
Does exposure to phthalate alternatives affect inflammation in children as measured by alpha-1-acid glycoprotein?
We will investigate this question using NHANES data from 2017-2018 of the serum inflammatory biomarker alpha-1-acid glycoprotein and urinary phthalate alternative metabolite levels, along with other relevant data (urinary creatinine levels and sociodemographic data).

# reading in relevant NHANES 2017-2018 data
* NOTE: phthalate and AGP data have special sample weights we need to account for in analysis
* weighting module: https://wwwn.cdc.gov/nchs/nhanes/tutorials/Weighting.aspx
```{r}
# phthalates and plasticizers metabolites - urine 
# codebook, LODs: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/PHTHTE_J.htm
phthte_urine =
  read.xport("data/PHTHTE_J.XPT") %>% 
  janitor::clean_names()

# urine creatinine levels 
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/ALB_CR_J.htm
creatinine_urine =
  read.xport("data/ALB_CR_J.XPT") %>% 
  janitor::clean_names()

# Alpha-1-Acid Glycoprotein - Serum
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/SSAGP_J.htm
agp_serum =
  read.xport("data/SSAGP_J.XPT") %>% 
  janitor::clean_names()

# Demographic variables and sample weights 
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DEMO_J.htm
demo =
  read.xport("data/DEMO_J.XPT")
```

# cleaning variable names of phthalate urine data 
```{r}
phthte_urine =
  phthte_urine %>% 
  rename(weights = wtsb2yr) %>%
  rename_with(~ gsub("^urx", "", .), starts_with("urx"))
  
# for now, leaving the <LOD comment columns (i.e, columns started wit urd) as they are 
# 2986 obs
```

# merging phthalate urine data with urinary creatinine levels and then adjusting concentrations for creatinine
```{r}
# dropping albumin from creatinine dataset
creatinine_urine =
  creatinine_urine %>% 
  select(seqn, urxucr, urducrlc) #dropping urxcrs bc i dont want creatinine expressed in those units
#7936 obs

#merging creatinine data with phthalate data
phthte_creatinine =
  merge(creatinine_urine, phthte_urine, by = "seqn") %>%
  select(!starts_with("urd")) %>% #dropped <LOD columns for ease of pivoting
  pivot_longer(cols = cnp:mzp,  
               names_to = "analyte_code",
               values_to = "reported_result") #pivoted longer so that i can map in the next step

#56734 obs 

#urinary [phth] in ng/mL
# urinary [creatinine] in mg/dL so need to get this to ng/mL

# Urinary creatinine concentrations, specific gravity, and osmolality are common methods for adjusting dilution. The most widely used method is creatinine adjustment that involves dividing the analyte concentration by the creatinine concentration. Analyte results are then reported as weight of analyte per gram of creatinine (micrograms analyte per gram creatinine).

# making a new column with creatinine-adjusted [phthalate]
phthte_creatinine =
  phthte_creatinine %>% 
  mutate(ng_per_ml_creatinine = urxucr*10000) %>% #here converting mg/dL to ng/mL--> 1mg/dL = 10,000ng/mL
  mutate(adj_conc = reported_result/ng_per_ml_creatinine)

#visualizing adjusted concentrations but on the log scale bc the numbers are now so small lol ... i wonder if im just being dumb or if NHANES really doesnt report specific gravity data 
boxplot(log10(phthte_creatinine$adj_conc))
hist(log10(phthte_creatinine$adj_conc))
```



