---
title: "nonlinear_regression"
output: html_document
date: "2024-03-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(foreign)
library(dplyr)
library(mgcv)
library(splines)
```

### Reading in Relevant NHANES 2017-2018 Data
* NOTE: phthalate and AGP data have special sample weights we need to account for in analysis
* weighting module: https://wwwn.cdc.gov/nchs/nhanes/tutorials/Weighting.aspx
```{r message=FALSE, warning=FALSE}
# phthalates and plasticizers metabolites - urine 
# codebook, LODs: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/PHTHTE_J.htm
phthte_urine =
  read.xport("data/PHTHTE_J.XPT") %>% 
  janitor::clean_names()

# urine creatinine levels 
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/ALB_CR_J.htm
creatinine_urine =
  read.xport("data/ALB_CR_J.XPT") %>% 
  janitor::clean_names()

# Alpha-1-Acid Glycoprotein - Serum
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/SSAGP_J.htm
agp_serum =
  read.xport("data/SSAGP_J.XPT") %>% 
  janitor::clean_names()

# Demographic variables and sample weights 
# codebook: https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DEMO_J.htm
demo =
  read.xport("data/DEMO_J.XPT") |> 
  janitor::clean_names()
```

### Cleaning Variable Names of Phthalate Urine Data 
```{r message=FALSE, warning=FALSE}
phthte_urine =
  phthte_urine %>% 
  rename(weights = wtsb2yr) %>%
  rename_with(~ gsub("^urx", "", .), starts_with("urx")) %>% 
  filter(!weights == "0")
```

### Adjusting Phthalate Urine Data for Creatinine
```{r message=FALSE, warning=FALSE}
# merging Phthalate urine data with urinary creatinine levels and then adjusting concentrations for creatinine
# dropping albumin from creatinine dataset
creatinine_urine =
  creatinine_urine %>% 
  select(seqn, urxucr, urducrlc) #dropping urxcrs bc i dont want creatinine expressed in those units
#7936 obs

#merging creatinine data with phthalate data
phthte_creatinine =
  merge(creatinine_urine, phthte_urine, by = "seqn") %>%
  select(!starts_with("urd")) %>% #dropped <LOD columns for ease of pivoting
  pivot_longer(cols = cnp:mzp,  
               names_to = "analyte_code",
               values_to = "reported_result") #pivoted longer so that i can map in the next step

#54169 obs 

#urinary [phth] in ng/mL
# urinary [creatinine] in mg/dL so need to get this to ng/mL

# Urinary creatinine concentrations, specific gravity, and osmolality are common methods for adjusting dilution. The most widely used method is creatinine adjustment that involves dividing the analyte concentration by the creatinine concentration. Analyte results are then reported as weight of analyte per gram of creatinine (micrograms analyte per gram creatinine).

# making a new column with creatinine-adjusted [phthalate]
phthte_creatinine =
  phthte_creatinine %>% 
  mutate(ng_per_ml_creatinine = urxucr*10000) %>% #here converting mg/dL to ng/mL--> 1mg/dL = 10,000ng/mL
  mutate(adj_conc = reported_result/ng_per_ml_creatinine)
```

### Cleaning Variable Names of Demographic Data
```{r message=FALSE, warning=FALSE}
# drop unneeded variables from dataset 
demo <- subset(demo, select = -c(sddsrvyr, ridstatr, ridexprg, dmdmartl, dmdeduc2, dmqmiliz, dmqadfc, dmdhrgnd, dmdhragz, dmdhredz, dmdhrmaz, dmdhsedz))

demo <- demo |> 
  rename(gender = riagendr, 
         age_yrs_screen = ridageyr, 
         age_mon_screen = ridagemn, 
         race_ethn1 = ridreth1,
         race_ethn3 = ridreth3, 
         time_period = ridexmon, 
         age_mon_screen = ridagemn,
         age_mon_exam = ridexagm, 
         birth_country = dmdborn4, 
         citizenship = dmdcitzn,
         years_us = dmdyrsus, 
         education = dmdeduc3, 
         lang_pers_quest = sialang, 
         proxy_pers_quest = siaproxy, 
         interpret_pers_quest = siaintrp, 
         lang_fam_interview = fialang, 
         proxy_fam_interview = fiaproxy, 
         interpret_fam_interview = fiaintrp,
         lang_capi = mialang, 
         proxy_capi = miaproxy, 
         interpret_capi = miaintrp, 
         lang_acasi = aialanga, 
         hh_size = dmdhhsiz, 
         fam_size = dmdfmsiz, 
         hh_young_children = dmdhhsza,
         hh_children = dmdhhszb, 
         hh_adults = dmdhhsze, 
         weight_2yr_intervew = wtint2yr,
         weight_2yr_mec = wtmec2yr, 
         variance_psu = sdmvpsu, 
         variance_stratum = sdmvstra, 
         income_hh_annual = indhhin2, 
         income_family_annual = indfmin2, 
         income_poverty_ratio = indfmpir)
```

### Subsetting Demographic Data to Ages 3-16 
```{r message=FALSE, warning=FALSE}
demo <- demo |> 
  filter(age_yrs_screen >= 3 & age_yrs_screen <= 16)
```

### Joining demographic, phthalate, and agp data
```{r}
#making phthalate adjusted with creatinine dataset wide so can join with demo data
wide_phth_creat =
  phthte_creatinine %>%
  select(-reported_result, -ng_per_ml_creatinine) %>% 
  pivot_wider(names_from = "analyte_code",
              values_from = "adj_conc") 
#noticed that participant #93739, 93771, has no data for any of the pththalate metabolites, even in phthte_urine dataset, so should drop them from further analysis?
# there are many participants with no values for any of them , even in phthte_urine dataset 
# went back to checkin_feb_20_deliverable to get rid of them there 


all_data =
  inner_join(demo, wide_phth_creat, by = "seqn") #915 people with demo and phthalate data


exp_outcome =
  inner_join(wide_phth_creat, agp_serum, by = "seqn") %>% 
  filter(!is.na(ssagp)) #662 participants that provided agp samples and have phthalate data for 
  
summary(exp_outcome)
(292/954)*100
```

#Non-linear exposure-response curves 
```{r}
#metabolites: cnp, cop, ecp, ecpt, hipb, mbp, mc1, mcoh, mep, mhbp, mhh, mhht, mhnc, mhp, mib, mnp, moh, monp, mzp

#first fit linear model 

lin_model <- lm(ssagp ~ cnp + cop + ecp + ecpt + hibp + mbp + mc1 + mcoh + mep + mhbp + mhh + mhht + mhnc + mhp + mib + mnp + moh + monp + mzp, data = exp_outcome)

summary(lin_model)

#fit complete model with penalized splines 

total_model <- gam(ssagp  ~ s(cnp) + s(cop) + s(ecp) + s(ecpt) + s(hibp) + s(mbp) + s(mc1) + s(mcoh) + s(mep) + s(mhbp) + s(mhh) + s(mhht) + s(mhnc) + s(mhp) + s(mib) + s(mnp) + s(moh) + s(monp) + s(mzp), data = exp_outcome)

summary(total_model)
#edf not equal to 1 for ecpt, mcoh, mep, mhht, mhnc, mhp, mnp, mzp
#going to fit GAM models with penalized splines for those 8 metabolites
```

### Indidual penalized splines for each metabolite with edf > 1 

```{r ecpt model}
# Fit GAM model for ecpt with penalized spline 
ecpt_model <- gam(ssagp ~ s(ecpt), data = exp_outcome)

# Generate predictions and standard errors
pred_ecpt <- predict(ecpt_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_ecpt_df <- data.frame(
  pred_centered = pred_ecpt$fit,
  se = pred_ecpt$se.fit,
  lci_centered = pred_ecpt$fit - 1.96 * pred_ecpt$se.fit,
  uci_centered = pred_ecpt$fit + 1.96 * pred_ecpt$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_ecpt_df <- cbind(exp_outcome, pred_ecpt_df)|> 
  mutate(metabolite_model = "ecpt")

# Uncenter data
pred_ecpt_df$pred_agp <- pred_ecpt_df$pred_centered + mean(exp_outcome$ssagp)
pred_ecpt_df$lci_agp <- pred_ecpt_df$lci_centered + mean(exp_outcome$ssagp)
pred_ecpt_df$uci_agp <- pred_ecpt_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_ecpt_df, aes(x = ecpt, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("ecpt (ug/L)") + 
  ylab("AGP (mg/mL)") + 
ggtitle("Non-linear relationship with penalized splines between ecpt and AGP")
```

```{r mcoh model}
# Fit GAM model for ecpt with penalized spline 
mcoh_model <- gam(ssagp ~ s(mcoh), data = exp_outcome)

# Generate predictions and standard errors
pred_mcoh <- predict(mcoh_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mcoh_df <- data.frame(
  pred_centered = pred_mcoh$fit,
  se = pred_mcoh$se.fit,
  lci_centered = pred_mcoh$fit - 1.96 * pred_mcoh$se.fit,
  uci_centered = pred_mcoh$fit + 1.96 * pred_mcoh$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mcoh_df <- cbind(exp_outcome, pred_mcoh_df)|> 
  mutate(metabolite_model = "mcoh")

# Uncenter data
pred_mcoh_df$pred_agp <- pred_mcoh_df$pred_centered + mean(exp_outcome$ssagp)
pred_mcoh_df$lci_agp <- pred_mcoh_df$lci_centered + mean(exp_outcome$ssagp)
pred_mcoh_df$uci_agp <- pred_mcoh_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mcoh_df, aes(x = mcoh, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mcoh (ug/L)") + 
  ylab("AGP (mg/mL)") + 
ggtitle("Non-linear relationship with penalized splines between mcoh and AGP")
```

```{r mep model}
# Fit GAM model for ecpt with penalized spline 
mep_model <- gam(ssagp ~ s(mep), data = exp_outcome)

# Generate predictions and standard errors
pred_mep <- predict(mep_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mep_df <- data.frame(
  pred_centered = pred_mep$fit,
  se = pred_mep$se.fit,
  lci_centered = pred_mep$fit - 1.96 * pred_mep$se.fit,
  uci_centered = pred_mep$fit + 1.96 * pred_mep$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mep_df <- cbind(exp_outcome, pred_mep_df) |> 
  mutate(metabolite_model = "mep")

# Uncenter data
pred_mep_df$pred_agp <- pred_mep_df$pred_centered + mean(exp_outcome$ssagp)
pred_mep_df$lci_agp <- pred_mep_df$lci_centered + mean(exp_outcome$ssagp)
pred_mep_df$uci_agp <- pred_mep_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mep_df, aes(x = mep, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mep (ug/L)") + 
  ylab("AGP (mg/mL)") + 
ggtitle("Non-linear relationship with penalized splines between mep and AGP")
```

The mhht model below is likely not showing any reportable data due to the increasingly wide confidence interval at higher levels of metabolite concentration
```{r mhht model}
# Fit GAM model for ecpt with penalized spline 
mhht_model <- gam(ssagp ~ s(mhht), data = exp_outcome)

# Generate predictions and standard errors
pred_mhht <- predict(mhht_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mhht_df <- data.frame(
  pred_centered = pred_mhht$fit,
  se = pred_mhht$se.fit,
  lci_centered = pred_mhht$fit - 1.96 * pred_mhht$se.fit,
  uci_centered = pred_mhht$fit + 1.96 * pred_mhht$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mhht_df <- cbind(exp_outcome, pred_mhht_df) |> 
  mutate(metabolite_model = "mhht")

# Uncenter data
pred_mhht_df$pred_agp <- pred_mhht_df$pred_centered + mean(exp_outcome$ssagp)
pred_mhht_df$lci_agp <- pred_mhht_df$lci_centered + mean(exp_outcome$ssagp)
pred_mhht_df$uci_agp <- pred_mhht_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mhht_df, aes(x = mhht, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mhht (ug/L)") + 
  ylab("AGP (mg/mL)") + 
  ggtitle("Non-linear relationship with penalized splines between mhht and AGP")
```

```{r mhnc model}
# Fit GAM model for ecpt with penalized spline 
mhnc_model <- gam(ssagp ~ s(mhnc), data = exp_outcome)

# Generate predictions and standard errors
pred_mhnc <- predict(mhnc_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mhnc_df <- data.frame(
  pred_centered = pred_mhnc$fit,
  se = pred_mhnc$se.fit,
  lci_centered = pred_mhnc$fit - 1.96 * pred_mhnc$se.fit,
  uci_centered = pred_mhnc$fit + 1.96 * pred_mhnc$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mhnc_df <- cbind(exp_outcome, pred_mhnc_df)|> 
  mutate(metabolite_model = "mhnc")

# Uncenter data
pred_mhnc_df$pred_agp <- pred_mhnc_df$pred_centered + mean(exp_outcome$ssagp)
pred_mhnc_df$lci_agp <- pred_mhnc_df$lci_centered + mean(exp_outcome$ssagp)
pred_mhnc_df$uci_agp <- pred_mhnc_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mhnc_df, aes(x = mhnc, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mhnc (ug/L)") + 
  ylab("AGP (mg/mL)") + 
  ggtitle("Non-linear relationship with penalized splines between mhnc and AGP")
```

This one also has an extremely wide confidence interval at larger values
```{r mhp model}
# Fit GAM model for ecpt with penalized spline 
mhp_model <- gam(ssagp ~ s(mhp), data = exp_outcome)

# Generate predictions and standard errors
pred_mhp <- predict(mhp_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mhp_df <- data.frame(
  pred_centered = pred_mhp$fit,
  se = pred_mhp$se.fit,
  lci_centered = pred_mhp$fit - 1.96 * pred_mhp$se.fit,
  uci_centered = pred_mhp$fit + 1.96 * pred_mhp$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mhp_df <- cbind(exp_outcome, pred_mhp_df) |> 
  mutate(metabolite_model = "mhp")

# Uncenter data
pred_mhp_df$pred_agp <- pred_mhp_df$pred_centered + mean(exp_outcome$ssagp)
pred_mhp_df$lci_agp <- pred_mhp_df$lci_centered + mean(exp_outcome$ssagp)
pred_mhp_df$uci_agp <- pred_mhp_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mhp_df, aes(x =mhp, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mhp (ug/L)") + 
  ylab("AGP (mg/mL)") + 
  ggtitle("Non-linear relationship with penalized splines between mhp and AGP")
```

```{r mnp model}
# Fit GAM model for ecpt with penalized spline 
mnp_model <- gam(ssagp ~ s(mnp), data = exp_outcome)

# Generate predictions and standard errors
pred_mnp <- predict(mnp_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mnp_df <- data.frame(
  pred_centered = pred_mnp$fit,
  se = pred_mnp$se.fit,
  lci_centered = pred_mnp$fit - 1.96 * pred_mnp$se.fit,
  uci_centered = pred_mnp$fit + 1.96 * pred_mnp$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mnp_df <- cbind(exp_outcome, pred_mnp_df) |> 
  mutate(metabolite_model = "mnp")

# Uncenter data
pred_mnp_df$pred_agp <- pred_mnp_df$pred_centered + mean(exp_outcome$ssagp)
pred_mnp_df$lci_agp <- pred_mnp_df$lci_centered + mean(exp_outcome$ssagp)
pred_mnp_df$uci_agp <- pred_mnp_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mnp_df, aes(x =mnp, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mnp (ug/L)") + 
  ylab("AGP (mg/mL)") + 
  ggtitle("Non-linear relationship with penalized splines between mnp and AGP")
```

Another situation with very wide confidence intervals at higher metabolite concentrations
```{r mzp model}
# Fit GAM model for ecpt with penalized spline 
mzp_model <- gam(ssagp ~ s(mzp), data = exp_outcome)

# Generate predictions and standard errors
pred_mzp <- predict(mzp_model, se.fit = TRUE)

# Create dataframe for predictions and standard errors
pred_mzp_df <- data.frame(
  pred_centered = pred_mzp$fit,
  se = pred_mzp$se.fit,
  lci_centered = pred_mzp$fit - 1.96 * pred_mzp$se.fit,
  uci_centered = pred_mzp$fit + 1.96 * pred_mzp$se.fit
)

#exclude missing values from original dataset and bind two datasets 
exp_outcome <- exp_outcome[complete.cases(exp_outcome), ]

# Combine predictions and original dataset
pred_mzp_df <- cbind(exp_outcome, pred_mzp_df) |> 
  mutate(metabolite_model = "mzp")

# Uncenter data
pred_mzp_df$pred_agp <- pred_mzp_df$pred_centered + mean(exp_outcome$ssagp)
pred_mzp_df$lci_agp <- pred_mzp_df$lci_centered + mean(exp_outcome$ssagp)
pred_mzp_df$uci_agp <- pred_mzp_df$uci_centered + mean(exp_outcome$ssagp)

# Plot
ggplot(pred_mzp_df, aes(x =mzp, y = pred_agp)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lci_agp, ymax = uci_agp), fill = "grey", alpha = 0.5) + 
  xlab("mzp (ug/L)") + 
  ylab("AGP (mg/mL)") + 
  ggtitle("Non-linear relationship with penalized splines between mzp and AGP")
```

```{r Demographic Covariate Model, linear regression}

#Joining current demographic values we have with the exp_outcome table
all_exp_outcome = inner_join(exp_outcome, demo, by = "seqn") |> 
  mutate(gender_female = if_else(gender == 1, 1, 0)) |>  #making a binary assignment of gender for multiple regression male = 0, female = 1
  mutate(eth_mex = case_when(race_ethn3 == 1 ~ 1, TRUE ~ 0),
         eth_hisp_oth = case_when(race_ethn3 == 2 ~ 1, TRUE ~ 0),
         eth_white = case_when(race_ethn3 == 3 ~ 1, TRUE ~ 0),
         eth_black = case_when(race_ethn3 == 4 ~ 1, TRUE ~ 0),
         eth_asian = case_when(race_ethn3 == 6 ~ 1, TRUE ~ 0),
         eth_nonhispanic_multirace = case_when(race_ethn3 == 7 ~ 1, TRUE ~ 0)) 

#Associations of demo variables alone
demographic_model = lm(ssagp ~ gender_female + age_yrs_screen + income_poverty_ratio + eth_mex + eth_hisp_oth + eth_white + eth_black + eth_asian + eth_nonhispanic_multirace, data = all_exp_outcome)

summary(demographic_model)

#Adding covariates to simple multiple regression 
lin_covari_model = lm(ssagp ~ cnp + cop + ecp + ecpt + hibp + mbp + mc1 + mcoh + mep + mhbp + mhh + mhht + mhnc + mhp + mib + mnp + moh + monp + mzp + gender_female + age_yrs_screen + income_poverty_ratio + eth_mex + eth_hisp_oth + eth_white + eth_black + eth_asian + eth_nonhispanic_multirace, data = all_exp_outcome)

summary(lin_covari_model)
```

Using this initial list of covariates, we do not currently have enough evidence to suggest a statistically significant difference in ssagp differing across ethnic groups, gender, or income poverty ratio. Under these adjusted assumptions, both mcoh and mhp are the metabolites tht have the most significant impact on ssagp.

**Returning to Total Model with Penalized Splines, adding covariates above**
```{r Adding Covariates to Penalized Spline Regression using GAM}
total_covari_model <- gam(ssagp  ~ s(cnp) + s(cop) + s(ecp) + s(ecpt) + s(hibp) + s(mbp) + s(mc1) + s(mcoh) + s(mep) + s(mhbp) + s(mhh) + s(mhht) + s(mhnc) + s(mhp) + s(mib) + s(mnp) + s(moh) + s(monp) + s(mzp)+ gender_female + age_yrs_screen + income_poverty_ratio + eth_mex + eth_hisp_oth + eth_white + eth_black + eth_asian + eth_nonhispanic_multirace, data = all_exp_outcome)

summary(total_covari_model)
#when adjusting for covariates, ecpt, mcoh, mhh, mhht, mib are the six metabolites that have an edf greater than 1
```
In this model, there are greater differences between children's ethnicities and ssagp concentration, where Mexican American and Black children and to a lesser extent white children are having a significant per unit increase in ssagp concentration for adjusted phthalate metabolite exposure.
